### 2021-11-01 docs/manuscript/manuscript.tex
Small improvements and fix typos

 \item {\bf Mutate GRNs randomly and evaluate fitness again.} Three types
 of mutations were possible in our algorithm:
1) changes in the numerical value of the degradation parameter, 2) changes
in the numerical value of the diffusion parameter and 3) changes in the
numerical values of the interaction strengths. The type of mutation to be
performed was selected randomly with a probability of 0.15 for type 1 and
type 2 mutations and 0.7 for type 3 mutations. If type 3 mutation was
selected, either the interaction strength was set to zero (with a
probability of 0.2) or was changed randomly (with a probability of 0.8).
If fitness of the mutated GRN was greater or equal than that of the
previous GRN, mutated GRN was saved and the cycle continued from step 2.
This cycle was repeated 50000 times, and a GRN sampled at a given step
and with a fitness score ≥ 0.95 was saved to a file as
long as its associated topology had not been sampled in previous steps.

@@

In order to group all the isomorphic (i.e. topologically equivalent) networks from
the totality (2061) of sampled GRNs we used a
classification algorithm that selected a GRN and generated all
isomorphs of that GRN by interchanging rows of the adjacency matrix
($i \leftrightarrow j$) while interchanging the corresponding columns
($i+1 \leftrightarrow j+1$). This kind of transformation of the adjacency
matrix results in isospectral graphs that are also isomorphic to the
original graph and is comparable to generate a permutation matrix $P$ and
multiplying $PMP^{T}$, where $M$ is the adjacency matrix and $P^{T}$ the
transpose of $P$. The algorithm then compared these isomorphs with another
GRN and classified the topologies in the same group if at least one of the
transformed adjacency matrices (or the original one) was identical to that
of the other GRN, i. e., if they were isomorphic networks.

@@

\subsection*{Pattern formation under varying size of the morphogenetic field}
We evaluated the ability of each of the sampled GRNs to generate a striped
pattern of gene expression under varying sizes of the morphogenetic field.
Specifically, we calculated the fitness of each GRN across morphogenetic fields
composed of 10, 20, 40, and 50 cells. In these simulations we redefined the
morphogen gradient in such a way that the initial and final concentration were
constant in all the fields (\nameref{S3_Fig}A). In order to know if topologies
with the same topology had a similar fitness in the different fields, we performed
a Principal Component Analysis in Scikit-learn~0.24.2~\cite{sklearn} using the
fitness values of each in the four different fields (see \nameref{S3_Fig}).

@@

\begin{figure}[!h]
%  \includegraphics[width=\textwidth]{figures/results/Fig2}
 \caption{\bf Distribution of abundances of network topologies.}
 Some of the most frequent network topologies shown in the grid
 contain the I3-FFL network motif (shown on the left). In this network
 motif the gene B is activated in an indirect way by the morphogen through
 the activation of A (positive interactions), and at the same time, it
 is inactivated in a direct way by the morphogen (negative interaction).
 Topologies are numbered in order of decreasing abundance, being the
 Topology 1 the most abundant with a number of 95 genotypes. The number of
 genotypes belonging to each network topology is presented below
 topology number.
 Note that network topology number 9 produces the striped phenotype using a
 two-node GRN.
 \label{fig:distopol}
\end{figure}

@@

The neutral network (a metagraph in which each node is a network topology)
shows that most of the network topologies are grouped into a connected graph.
This graph is composed of 639 nodes that are accessible by one mutational step.
The 15 most abundant network topologies, with the exception of
number 10, are part of this graph (Fig~\ref{fig:neutral-network}). We also
found nodes inside the neutral network that had more connections between them
than with other nodes, which leads to further clustering of nodes. Interestingly in
almost every cluster one can find one or more of the 15 most abundant network
topologies. In addition, we found that the first eight most abundant network
topologies were located in the largest cluster.

@@

The degree distribution of this network seems to follow a Poisson distribution
(Pearson’s $\chi^2 = 12.8902$, $\text{p-value} = 0.115684$) characteristic of a
Erdös-Rényi network with many nodes and edges~\cite{Erdos1959}. This kind of
network presents a great
number of nodes with low degree and a few nodes with high degree
(Fig~\ref{fig:deg-dist}A). Nodes corresponding to topologies number one and
three showed the highest degree ($deg(1) = deg (3) = 14$). Curiously, we found
that nodes with higher degree turned out to be also those network topologies
with higher abundance
(Fig~\ref{fig:deg-dist}B). This relationship between abundance and node degree
was corroborated by the Spearman's rank correlation coefficient
($r_S = 0.353752$, $p = 6.2933\times10^{-23} $).

@@

\begin{figure}
%  \includegraphics[width=\textwidth]{figures/results/Fig5}
 \caption{\bf Network of topologies separated by two mutational steps.}
 The network shows that almost all the topologies are connected by two or less
 mutational steps. Big red circle represents topology 420 that remains
 unconnected from the main graph. Big black circles correspond to the 15
 most abundant topologies described in Fig~\ref{fig:neutral-network}.
 Different colors in the graph represent different clusters of topologies.
 \label{fig:2neut-net}
\end{figure}

@@

This is a complex mechanism reminiscent of the Gierer-Meinhardt model, which
to the best of our knowledge is reported for the first time in the context of
multiple morphogen inputs, and could thus be of particular interest to for
synthetic biology implementations to assess the feasibility and robustness
conferred by this mechanism to GRNs.

@@

In contrast to this robustness to
qualitative and quantitative changes, we found that many GRNs were susceptible
to changes in morphogenetic field size. However, interestingly, GRNs with the same
topology behave in different ways for each field size~(\nameref{S3_Fig}C),
indicating that the specific values of network parameters are more important
than the network topology when it comes to producing the same pattern at
different scales.

@@

However, as our model was less stringent and allowed the morphogen input to
interact with any gene, we found a significantly larger number of pattern-forming
topologies, which can operate under disparate combinations of diffusion and degradation
parameters~(\nameref{S7_Fig}). For example, in a single input GRN model, the I4-FFL
has not been found capable of generating a striped gene expression
pattern~\cite{munteanu_2014}, whereas, in our experimental setup, 20.17\% of
the topologies were found  to implement such a regulatory motif~(\nameref{S1_Table}).

@@

We also found topologies with a similar topology to that of the “opposing
gradients” reported by Schaerli et al.~\cite{Schaerli2018,Schaerli2014}.
Although the “opposing gradients” mechanism requires the constitutive expression
of two of the genes in the GRN, and our model does not take constitutive
expression into account, we observed that the GRNs that implement the “opposing
gradients” mechanism also tend to implement auto-regulation interactions on
these two nodes, bypassing in this way the lack of constitutive
expression as found in previous studies~\cite{munteanu_2014}.

@@

\section*{Acknowledgements}
The computational resources (Stevin Supercomputer Infrastructure) and
services used in this work were provided by the VSC (Flemish Supercomputer
Center), funded by Ghent University, FWO and the Flemish Government–department EWI.
All authors acknowledge partial support from CoDI-Universidad de Antioquia under
project ``Estudio de la Evolución de Circuitos de Regulación del Desarollo
Embrionario a través de la Biología Evolutiva de Sistemas'' code 2017-14367. All
authors acknowledge the Colombian youth for their defense of a better country.

>>>>

 \item {\bf Mutate GRNs randomly and evaluate fitness again.} Three types
 of mutations were possible in our algorithm:
1) changes in the numerical value of the degradation parameter, 2) changes
in the numerical value of the diffusion parameter and 3) changes in the
numerical values of the interaction strengths. The type of mutation to be
performed was selected randomly with a probability of 0.15 for type 1 and
type 2 mutations and 0.7 for type 3 mutations. If type 3 mutation was
selected, either the interaction strength was set to zero (with a
probability of 0.2) or was changed randomly (with a probability of 0.8).
If fitness of the mutated GRN was greater or equal than that of the
previous GRN, the mutated GRN was saved and the cycle continued from step 2.
This cycle was repeated 50000 times, and a GRN sampled at a given step
and with a fitness score ≥ 0.95 was saved to a file as
long as its associated topology had not been sampled in previous steps.

@@

In order to group all the isomorphic (i.e. topologically equivalent) networks from
the totality (2061) of sampled GRNs we used a
classification algorithm that selected a GRN and generated all
isomorphs of that GRN by interchanging rows of the adjacency matrix
($i \leftrightarrow j$) while interchanging the corresponding columns
($i+1 \leftrightarrow j+1$). This kind of transformation of the adjacency
matrix results in isospectral graphs that are also isomorphic to the
original graph and is equivalent to generate a permutation matrix $P$ and
multiplying $PMP^{T}$, where $M$ is the adjacency matrix and $P^{T}$ the
transpose of $P$. The algorithm then compared these isomorphs with another
GRN and classified the topologies in the same group if at least one of the
transformed adjacency matrices (or the original one) was identical to that
of the other GRN, i. e., if they were isomorphic networks.

@@

\subsection*{Pattern formation under varying size of the morphogenetic field}
We evaluated the ability of each of the sampled GRNs to generate a striped
pattern of gene expression under varying sizes of the morphogenetic field.
Specifically, we calculated the fitness of each GRN across morphogenetic fields
composed of 10, 20, 40, and 50 cells. In these simulations we redefined the
morphogen gradient in such a way that the initial and final concentration were
constant in all the fields (\nameref{S3_Fig}A). In order to know if GRNs
with the same topology had a similar fitness in the different fields, we performed
a Principal Component Analysis in Scikit-learn~0.24.2~\cite{sklearn} using the
fitness values of each in the four different fields (see \nameref{S3_Fig}).

@@

\begin{figure}[!h]
%  \includegraphics[width=\textwidth]{figures/results/Fig2}
 \caption{\bf Distribution of abundances of network topologies.}
 Some of the most frequent network topologies shown in the grid
 contain the I3-FFL network motif (shown on the left). In this network
 motif the gene B is activated in an indirect way by the morphogen through
 the activation of A (positive interactions), and at the same time, it
 is inactivated in a direct way by the morphogen (negative interaction).
 Topologies are numbered in order of decreasing abundance, being the
 Topology 1 the most abundant with a number of 95 genotypes. The number of
 genotypes belonging to each network topology is presented below the
 topology number.
 Note that network topology number 9 produces the striped phenotype using a
 two-node GRN.
 \label{fig:distopol}
\end{figure}

@@

The neutral network (a metagraph in which each node is a network topology)
shows that most of the network topologies are grouped into a connected graph.
This graph is composed of 639 nodes that are accessible by one mutational step.
The 15 most abundant network topologies, with the exception of
number 10, are part of this graph (Fig~\ref{fig:neutral-network}). We also
found nodes inside the neutral network that had more connections between them
than with other nodes, which leads to further clustering of nodes. Interestingly, in
almost every cluster one can find one or more of the 15 most abundant network
topologies. In addition, we found that the first eight most abundant network
topologies were located in the largest cluster.

@@

The degree distribution of this neutral network seems to follow a Poisson distribution
(Pearson’s $\chi^2 = 12.8902$, $\text{p-value} = 0.115684$) characteristic of a
Erdös-Rényi network with many nodes and edges~\cite{Erdos1959}. This kind of
network presents a great
number of nodes with low degree and a few nodes with high degree
(Fig~\ref{fig:deg-dist}A). Nodes corresponding to topologies number one and
three showed the highest degree ($deg(1) = deg (3) = 14$). Curiously, we found
that nodes with higher degree turned out to be also those network topologies
with higher abundance
(Fig~\ref{fig:deg-dist}B). This relationship between abundance and node degree
was corroborated by the Spearman's rank correlation coefficient
($r_S = 0.353752$, $p = 6.2933\times10^{-23} $).

@@

\begin{figure}
%  \includegraphics[width=\textwidth]{figures/results/Fig5}
 \caption{\bf Neutral network of topologies separated by two mutational
 steps.}
 The network shows that almost all the topologies are connected by two or less
 mutational steps. Big red circle represents topology 420 that remains
 unconnected from the main graph. Big black circles correspond to the 15
 most abundant topologies described in Fig~\ref{fig:neutral-network}.
 Different colors in the graph represent different clusters of topologies.
 \label{fig:2neut-net}
\end{figure}

@@

This is a complex mechanism reminiscent of the Gierer-Meinhardt model, which
to the best of our knowledge is reported for the first time in the context of
multiple morphogen inputs, and could thus be of particular interest for
synthetic biology implementations to assess the feasibility and robustness
conferred by this mechanism to GRNs.

@@

In contrast to this robustness to
qualitative and quantitative changes, we found that many GRNs were susceptible
to changes in morphogenetic field size. However, GRNs with the same
topology behave in different ways for each field size~(\nameref{S3_Fig}C),
indicating that the specific values of network parameters are more important
than the network topology when it comes to producing the same pattern at
different scales.

@@

Moreover, as our model was less stringent and allowed the morphogen input to
interact with any gene, we found a significantly larger number of pattern-forming
topologies, which can operate under disparate combinations of diffusion and degradation
parameters~(\nameref{S7_Fig}). For example, in a single input GRN model, the I4-FFL
has not been found capable of generating a striped gene expression
pattern~\cite{munteanu_2014}, whereas, in our experimental setup, 20.17\% of
the topologies were found  to implement such a regulatory motif~(\nameref{S1_Table}).

@@

We also found GRNs with a similar topology to that of the “opposing
gradients” reported by Schaerli et al.~\cite{Schaerli2018,Schaerli2014}.
Although the “opposing gradients” mechanism requires the constitutive expression
of two of the genes in the GRN, and our model does not take constitutive
expression into account, we observed that the GRNs that implement the “opposing
gradients” mechanism also tend to implement auto-regulation interactions on
these two nodes, bypassing in this way the lack of constitutive
expression as found in previous studies~\cite{munteanu_2014}.

@@

\section*{Acknowledgements}
The computational resources (Stevin Supercomputer Infrastructure) and
services used in this work were provided by the VSC (Flemish Supercomputer
Center), funded by Ghent University, FWO and the Flemish Government–department EWI.
All authors acknowledge partial support from CODI-Universidad de Antioquia under
project ``Estudio de la Evolución de Circuitos de Regulación del Desarollo
Embrionario a través de la Biología Evolutiva de Sistemas'' code 2017-14367. All
authors acknowledge the Colombian youth for their defense of a better country.
